---
title: "credit_exploration"
output: html_document
---
# libraries
```{r}
library(shiny)
library(tidyverse)
library(glue)
library(DT)
library(lubridate)
library(broom)
```

```{r}
credit_raw <- read_csv("../data/loans_full_schema.csv") 

credit <- credit_raw |>
  mutate(
    default = if_else(
      loan_status %in% c("Charged Off", "Default", "Late (31-120 days)"),
      1L, 0L
    )
  )
#This model predicts probability of non-completion, not realized default.
```
#cleaning the data

Dropping a few things that leak future information and are not valid predictors.
```{r}
credit <- credit |>
  select(
    -emp_title,
    -installment,   # derived from loan_amount + term + rate
    -interest_rate, # outcome of underwriting, not input
    -grade,
    -sub_grade,
    -initial_listing_status,
    -disbursement_method
  )
```

Employment Length is categorical and shoes be converted to numeric
```{r}
credit <- credit |>
  mutate(
    emp_length = as.numeric(emp_length),
    emp_length = if_else(is.na(emp_length), 0 , emp_length)
  )
```

Lets make some other variables human readable
```{r}
#state
credit <- credit |>
  mutate(state = factor(state))

#Homeownership
credit <- credit |>
  mutate(homeownership = factor(homeownership))

#verified income
credit <- credit |>
  mutate(verified_income = factor(verified_income))

#application type
credit <- credit |>
  mutate(application_type = factor(application_type))

#loan purpose
credit <- credit |>
  mutate(loan_purpose = factor(loan_purpose))

```

Now lets clean up joint and individual loans
```{r}
credit <- credit |>
  mutate(
    total_income = if_else(
      application_type == "joint" & !is.na(annual_income_joint),
      annual_income_joint,
      annual_income
    ),
    total_dti = if_else(
      application_type == "joint" & !is.na(debt_to_income_joint),
      debt_to_income_joint,
      debt_to_income
    )
  )

#drop the raw components
credit <- credit |>
  select(
    -annual_income,
    -annual_income_joint,
    -debt_to_income,
    -debt_to_income_joint,
    -verification_income_joint
  )
```

Fixing the credit years variable
```{r}
credit <- credit |>
  mutate(
    # Ensure earliest_credit_line is numeric year
    earliest_credit_line = as.numeric(as.character(earliest_credit_line)),

    # Convert issue_month to date
    issue_date = as.Date(paste0("01-", issue_month), format = "%d-%b-%Y"),

    # Extract issue year
    issue_year = as.integer(format(issue_date, "%Y")),

    # Compute credit history length
    credit_history_years = issue_year - earliest_credit_line,

    # Guardrails
    credit_history_years = pmax(credit_history_years, 0),
    credit_history_years = pmin(credit_history_years, 50)
  ) |>
  select(
    -issue_month,
    -issue_date,
    -earliest_credit_line
  )

```


Now I want to create interpretable ratios
```{r}
credit <- credit |>
  mutate(
    credit_utilization = total_credit_utilized / total_credit_limit,
    debit_utilization  = total_debit_limit / pmax(num_active_debit_accounts, 1)
  )
```

Now I have to deal with missing values
```{r}
credit <- credit |>
  mutate(
    months_since_last_delinq = if_else(
      is.na(months_since_last_delinq), 999, months_since_last_delinq
    ),
    months_since_90d_late = if_else(
      is.na(months_since_90d_late), 999, months_since_90d_late
    )
  )

```

# Logistic Regression
```{r}
credit_model <- credit |>
  filter(
    loan_status %in% c(
      "Fully Paid",
      "Charged Off",
      "Late (31-120 days)"
    )
  )
```

create a target variable
```{r}
credit_model <- credit_model |>
  mutate(
    default = if_else(
      loan_status %in% c("Charged Off", "Late (31-120 days)"),
      1L, 0L
    )
  )

#dropping loan status because we labeled it
credit_model <- credit_model |>
  select(-loan_status)
```

```{r}
credit_model |>
  count(default) |>
  mutate(pct = n / sum(n))

```
Default rate somewhere in the 10–25% range

```{r}
table(credit$loan_status)
```
I excluded current loans because PD models require observed outcomes. Current and early-stage delinquent loans haven’t resolved yet, so including them would bias the target. I restricted training to finalized outcomes, which is standard practice in retail credit risk.

For a baseline PD model, we want:
-Credit behavior
-Capacity
-Stability
-Minimal redundancy
-No post-decision leakage

I needed to make sure all variables are usable in a model.
```{r}
credit_model <- credit_model |>
  mutate(
    # Explicit numeric conversions
    total_income = as.numeric(total_income),
    total_dti = as.numeric(total_dti),
    loan_amount = as.numeric(loan_amount),
    credit_history_years = as.numeric(credit_history_years),
    delinq_2y = as.numeric(delinq_2y),
    months_since_last_delinq = as.numeric(months_since_last_delinq),
    months_since_90d_late = as.numeric(months_since_90d_late),
    num_historical_failed_to_pay = as.numeric(num_historical_failed_to_pay),
    inquiries_last_12m = as.numeric(inquiries_last_12m),
    credit_utilization = as.numeric(credit_utilization),
    total_credit_lines = as.numeric(total_credit_lines),
    open_credit_lines = as.numeric(open_credit_lines),
    emp_length = as.numeric(emp_length),

    # Explicit categorical casting
    term = factor(term),
    homeownership = factor(homeownership),
    application_type = factor(application_type)
  )

```


```{r}
credit_model_lr <- glm(
  default ~
    total_income +
    total_dti +
    loan_amount +
    term +
    credit_history_years +
    delinq_2y +
    months_since_last_delinq +
    months_since_90d_late +
    num_historical_failed_to_pay +
    inquiries_last_12m +
    credit_utilization +
    total_credit_lines +
    open_credit_lines +
    emp_length +
    homeownership +
    application_type,
  data = credit_model,
  family = binomial()
)


```


```{r}
coef_table <- tidy(credit_model_lr, conf.int = TRUE, exponentiate = TRUE)

coef_table |>
  arrange(desc(estimate))

```

```{r}
credit_model_lr_results <- credit_model |>
  mutate(
    pd = predict(credit_model_lr, type = "response")
  )
summary(credit_model_lr_results)
```












