---
title: "credit_exploration"
output: html_document
---
# libraries
```{r}
library(shiny)
library(tidyverse)
library(glue)
library(DT)
library(lubridate)
library(broom)
```

```{r}
credit_raw <- read_csv("data/loans_full_schema.csv") 

credit <- credit_raw |>
  mutate(
    default = if_else(
      loan_status %in% c("Charged Off", "Default", "Late (31-120 days)"),
      1L, 0L
    )
  )
#This model predicts probability of non-completion, not realized default.
```
#cleaning the data

Dropping a few things that leak future information and are not valid predictors.
```{r}
credit <- credit |>
  select(
    -emp_title,
    -installment,   # derived from loan_amount + term + rate
    -interest_rate, # outcome of underwriting, not input
    -grade,
    -sub_grade,
    -initial_listing_status,
    -disbursement_method
  )
```

Employment Length is categorical and should be converted to numeric
```{r}
credit <- credit |>
  mutate(
    emp_length = as.numeric(emp_length),
    emp_length = if_else(is.na(emp_length), 0 , emp_length)
  )
```

Lets make some other variables human readable
```{r}
#state
credit <- credit |>
  mutate(state = factor(state))

#Homeownership
credit <- credit |>
  mutate(homeownership = factor(homeownership))

#verified income
credit <- credit |>
  mutate(verified_income = factor(verified_income))

#application type
credit <- credit |>
  mutate(application_type = factor(application_type))

#loan purpose
credit <- credit |>
  mutate(loan_purpose = factor(loan_purpose))

```

Now lets clean up joint and individual loans
```{r}
credit <- credit |>
  mutate(
    total_income = if_else(
      application_type == "joint" & !is.na(annual_income_joint),
      annual_income_joint,
      annual_income
    ),
    total_dti = if_else(
      application_type == "joint" & !is.na(debt_to_income_joint),
      debt_to_income_joint,
      debt_to_income
    )
  )

#drop the raw components
credit <- credit |>
  select(
    -annual_income,
    -annual_income_joint,
    -debt_to_income,
    -debt_to_income_joint,
    -verification_income_joint
  )
```

Fixing the credit years variable
```{r}
credit <- credit |>
  mutate(
    # Ensure earliest_credit_line is numeric year
    earliest_credit_line = as.numeric(as.character(earliest_credit_line)),

    # Convert issue_month to date
    issue_date = as.Date(paste0("01-", issue_month), format = "%d-%b-%Y"),

    # Extract issue year
    issue_year = as.integer(format(issue_date, "%Y")),

    # Compute credit history length
    credit_history_years = issue_year - earliest_credit_line,

    # Guardrails
    credit_history_years = pmax(credit_history_years, 0),
    credit_history_years = pmin(credit_history_years, 50)
  ) |>
  select(
    -issue_month,
    -issue_date,
    -earliest_credit_line
  )

```


Now I want to create interpretable ratios
```{r}
credit <- credit |>
  mutate(
    credit_utilization = total_credit_utilized / total_credit_limit,
    debit_utilization  = total_debit_limit / pmax(num_active_debit_accounts, 1)
  )
```

Now I have to deal with missing values
```{r}
credit <- credit |>
  mutate(
    months_since_last_delinq = if_else(
      is.na(months_since_last_delinq), 999, months_since_last_delinq
    ),
    months_since_90d_late = if_else(
      is.na(months_since_90d_late), 999, months_since_90d_late
    )
  )

```

Test the AUC and ROC on original data
```{r}
set.seed(123)  

n <- nrow(credit)

train_index <- sample(seq_len(n), size = 0.75 * n)

train_data <- credit[train_index, ]
test_data  <- credit[-train_index, ]

```

#ROC and AUC on samples of data set

```{r}
credit_model_trained_lr <- glm(
  default ~ .,
  data = train_data,
  family = binomial()
)

```
```{r}
library(pROC)

#Generate Predictions on TEST DATA ONLY
test_data$pd <- predict(
  credit_model_trained_lr,
  newdata = test_data,
  type = "response"
)

#ROC Object
roc_obj <- roc(
  response = test_data$default,
  predictor = test_data$pd
)

auc_value <- auc(roc_obj)

#Optimal Cutoff (Youden Index) 
opt_coords <- coords(
  roc_obj,
  x = "best",
  best.method = "youden",
  ret = c("threshold", "sensitivity", "specificity")
)

opt_threshold  <- opt_coords["threshold"]
sensitivity    <- opt_coords["sensitivity"]
specificity    <- opt_coords["specificity"]

# Predicted Classes Using Optimal Threshold
test_data$pred_class <- ifelse(test_data$pd >= opt_threshold, 1, 0)

#confusion matrix
conf_mat <- table(
  Actual = test_data$default,
  Predicted = test_data$pred_class
)

accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
```

Plot ROC with All Key Info
```{r}
png(
  filename = "roc_test_plot.png",
  width = 1200,
  height = 900,
  res = 150
)

plot(
  roc_obj,
  col = "blue",
  lwd = 3,
  main = "ROC Curve (Test Set)"
)

legend(
  "bottomright",
  legend = c(
    paste0("AUC: ", round(auc_value, 3)),
    paste0("Threshold: ", round(opt_threshold, 3)),
    paste0("Sensitivity: ", round(sensitivity, 3)),
    paste0("Specificity: ", round(specificity, 3)),
    paste0("Accuracy: ", round(accuracy, 3))
  ),
  bty = "n"
)

abline(a = 0, b = 1, lty = 2, col = "gray")

dev.off()

```
```{r}
getwd()

```
Explanation: 
- ROC shows how well the model ranks defaults above non-defaults
- AUC ≈ probability a random bad loan scores worse than a random good loan

0.5 = useless
0.6–0.7 = weak but usable
0.7–0.8 = solid credit model
0.8+ = very strong (rare for consumer credit)

# Logistic Regression
```{r}
credit_model <- credit |>
  filter(
    loan_status %in% c(
      "Fully Paid",
      "Charged Off",
      "Late (31-120 days)"
    )
  )
```

create a target variable
```{r}
credit_model <- credit_model |>
  mutate(
    default = if_else(
      loan_status %in% c("Charged Off", "Late (31-120 days)"),
      1L, 0L
    )
  )

#dropping loan status because we labeled it
credit_model <- credit_model |>
  select(-loan_status)
```

```{r}
credit_model |>
  count(default) |>
  mutate(pct = n / sum(n))

```
Default rate somewhere in the 10–25% range

```{r}
table(credit$loan_status)
```
I excluded current loans because PD models require observed outcomes. Current and early-stage delinquent loans haven’t resolved yet, so including them would bias the target. I restricted training to finalized outcomes, which is standard practice in retail credit risk.

For a baseline PD model, we want:
-Credit behavior
-Capacity
-Stability
-Minimal redundancy
-No post-decision leakage

I needed to make sure all variables are usable in a model.
```{r}
credit_model <- credit_model |>
  mutate(
    total_income = as.numeric(total_income),
    total_dti = as.numeric(total_dti),
    loan_amount = as.numeric(loan_amount),
    credit_history_years = as.numeric(credit_history_years),
    delinq_2y = as.numeric(delinq_2y),
    months_since_last_delinq = as.numeric(months_since_last_delinq),
    months_since_90d_late = as.numeric(months_since_90d_late),
    num_historical_failed_to_pay = as.numeric(num_historical_failed_to_pay),
    inquiries_last_12m = as.numeric(inquiries_last_12m),
    credit_utilization = as.numeric(credit_utilization),
    total_credit_lines = as.numeric(total_credit_lines),
    open_credit_lines = as.numeric(open_credit_lines),
    emp_length = as.numeric(emp_length),

    # Explicit categorical casting
    term = factor(term),
    homeownership = factor(homeownership),
    application_type = factor(application_type)
  )

```

```{r}
saveRDS(credit_model, file = "credit_model_RDS")
```

```{r}
credit_model_lr <- glm(
  default ~
    total_income +
    total_dti +
    loan_amount +
    term +
    credit_history_years +
    delinq_2y +
    months_since_last_delinq +
    months_since_90d_late +
    num_historical_failed_to_pay +
    inquiries_last_12m +
    credit_utilization +
    total_credit_lines +
    open_credit_lines +
    emp_length +
    homeownership +
    application_type,
  data = credit_model,
  family = binomial()
)


```

```{r}
saveRDS(credit_model_lr, file = "credit_model_lr_RDS")
```

```{r}
coef_table <- tidy(credit_model_lr, conf.int = TRUE, exponentiate = TRUE)

coef_table |>
  arrange(desc(estimate))

```




```{r}
library(ggplot2)

pd_plot <- ggplot(credit_model_lr_results, aes(x = pd, fill = factor(default))) +
  geom_histogram(
    bins = 50,
    position = "identity",
    alpha = 0.5
  ) +
  scale_fill_manual(
    values = c(
      "0" = "#003fba",  # No Default
      "1" = "#706363"   # Default
    )
  ) +
  labs(
    title = "Predicted PD Distribution by Loan Outcome",
    x = "Predicted Probability of Default",
    fill = "Default"
  )
```


```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    pd_decile = ntile(pd, 10)
  )

credit_model_lr_results |>
  group_by(pd_decile) |>
  summarise(
    avg_pd = mean(pd),
    default_rate = mean(default),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(pd_decile))

```

The model demonstrates strong ranking ability, with an AUC of X. Default rates increase monotonically across PD deciles, indicating effective risk separation.

```{r}
ggsave(
  filename = "pd_distribution.png",
  plot = pd_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```


# Risk Bands

```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    risk_band = case_when(
      pd < 0.05 ~ "Low",
      pd < 0.15 ~ "Medium",
      TRUE ~ "High"
    ),
    risk_band = factor(risk_band, levels = c("Low", "Medium", "High"))
  )

```

default risk bands
```{r}
credit_model_lr_results |>
  group_by(risk_band) |>
  summarise(
    avg_pd = mean(pd),
    observed_default_rate = mean(default),
    n = n(),
    .groups = "drop"
  )

```

visual of risk bands
```{r}
ggplot(credit_model_lr_results, aes(x = risk_band, fill = factor(default))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c("0" = "#003fba",
               "1" = "#706363") 
  ) +
  labs(
    title = "Observed Default Rate by Risk Band",
    x = "Risk Band",
    y = "Proportion",
    fill = "Default"
  )

```
decision logic

```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    decision = case_when(
      risk_band == "Low" ~ "Approve",
      risk_band == "Medium" ~ "Review",
      TRUE ~ "Decline"
    )
  )
credit_model_lr_results
```
Summary: 
Predicted PDs were translated into risk bands aligned with hypothetical credit policy thresholds. These bands allow consistent decisioning and enable transparent communication of risk to non-technical stakeholders.

# Lets explain somethings here

```{r}
coef_table <- tidy(
  credit_model_lr,
  conf.int = TRUE
)

coef_table <- coef_table |>
  mutate(
    estimate = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high)
  )

coef_table |>
  arrange(desc(estimate))


```

```{r}
coef_table_clean <- coef_table |>
  filter(term != "(Intercept)") |>
  mutate(
    direction = if_else(estimate > 1, "Increases Risk", "Decreases Risk"),
    pct_change_odds = (estimate - 1) * 100
  ) |>
  arrange(desc(abs(pct_change_odds)))
coef_table_clean
```

```{r}
#work to make sure this is working for each variable to explain what is making the prediction change
X <- model.matrix(credit_model_lr)
betas <- coef(credit_model_lr)

contrib <- sweep(X, 2, betas, `*`)

explain_one <- function(row_id) {
  tibble(
    feature = colnames(contrib),
    contribution = contrib[row_id, ]
  ) |>
    filter(feature != "(Intercept)") |>
    arrange(desc(abs(contribution)))
}

explain_one(1)
```

```{r}
explain_one(1) |>
  mutate(
    impact = case_when(
      contribution > 0 ~ "Increases PD",
      TRUE ~ "Decreases PD"
    )
  )

```

High credit utilization and renting status increase PD, while longer credit history reduces PD.

```{r}
full_individual_report <- function(row_id, model, data,
                                   risk_thresholds = c(Low = 0.05, Medium = 0.15),
                                   top_n = 3) {
  
  borrower <- data[row_id, , drop = FALSE]
  
  # Predicted PD
  predicted_PD <- as.numeric(
    predict(model, newdata = borrower, type = "response")
  )
  
  # Risk band
  risk_band <- dplyr::case_when(
    predicted_PD < risk_thresholds["Low"] ~ "Low",
    predicted_PD < risk_thresholds["Medium"] ~ "Medium",
    TRUE ~ "High"
  )
  
  # Model matrix for ONE borrower
  X <- model.matrix(model, borrower)
  betas <- coef(model)
  
  # Align coefficients to matrix columns
  betas <- betas[colnames(X)]
  
  # Proper element-wise multiplication (single row)
  contrib_log_odds <- as.numeric(X[1, ] * betas)
  
  # Convert to probability contribution
  prob <- predicted_PD
  contrib_prob <- contrib_log_odds * prob * (1 - prob)
  
  top_contributors <- tibble::tibble(
    feature = colnames(X),
    value = as.numeric(X[1, ]),
    contribution_prob = contrib_prob,
    direction = ifelse(contrib_prob > 0, "Increases Risk", "Decreases Risk")
  ) |>
    dplyr::filter(feature != "(Intercept)") |>
    dplyr::arrange(desc(abs(contribution_prob))) |>
    dplyr::slice_head(n = top_n)
  
  # Marginal effects
  variable_effects <- tibble::tibble(
    feature = names(betas),
    beta = betas
  ) |>
    dplyr::filter(feature != "(Intercept)") |>
    dplyr::mutate(
      prob_change = beta * prob * (1 - prob)
    ) |>
    dplyr::select(feature, prob_change)
  
  return(list(
    row_id = row_id,
    predicted_PD = predicted_PD,
    risk_band = risk_band,
    top_contributors = top_contributors,
    variable_effects = variable_effects
  ))
}

```

```{r}
print_individual_report(1, credit_model_lr, credit_model_lr_results)

```


```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    decision_reason = case_when(
      risk_band == "High" ~ "High predicted default risk",
      risk_band == "Medium" ~ "Moderate risk — requires review",
      TRUE ~ "Low predicted default risk"
    )
  )

```
The model produces a Probability of Default, maps it to risk bands, and provides feature-level explanations for each decision, aligning with model risk management and fair lending expectations.


```{r}
all_vars <- names(coef(credit_model_lr))
all_vars
```
## Tibble for high risk factors

for logistic regression 
n=XB

impact of a variable
Impact(J) = B(J)(X(J)- X(J))

create an interactive table that generates impact of each variable based on the comparison with the average value and actual value.

```{r}
explain_table <- reactive({
  
  # Current borrower
  new_data <- borrower_data()
  
  # Model matrix for borrower
  X_new <- model.matrix(
    delete.response(terms(credit_model_lr)),
    new_data
  )
  
  # Model matrix for entire training data
  X_train <- model.matrix(
    delete.response(terms(credit_model_lr)),
    credit_model
  )
  
  # Average borrower (column means)
  X_avg <- colMeans(X_train)
  
  betas <- coef(credit_model_lr)
  
  results <- tibble(
    feature = colnames(X_new),
    impact = NA_real_
  )
  
  #FOR LOOP
  for (i in seq_along(results$feature)) {
    
    var_name <- results$feature[i]
    
    if (var_name %in% names(betas)) {
      
      borrower_value <- X_new[1, var_name]
      avg_value <- X_avg[var_name]
      beta_value <- betas[var_name]
      
      results$impact[i] <- beta_value * (borrower_value - avg_value)
    }
  }
  
  results |>
    filter(feature != "(Intercept)") |>
    arrange(desc(abs(impact))) |>
    slice_head(n = 5)
})

```


# Overall Scaling

```{r}
library(dplyr)

# Identify numeric predictors used in the model
numeric_vars <- c(
  "total_income",
  "total_dti",
  "loan_amount",
  "credit_history_years",
  "delinq_2y",
  "months_since_last_delinq",
  "months_since_90d_late",
  "num_historical_failed_to_pay",
  "inquiries_last_12m",
  "credit_utilization",
  "total_credit_lines",
  "open_credit_lines",
  "emp_length"
)

# Scale numeric predictors
credit_model_scaled <- credit_model |>
  mutate(across(all_of(numeric_vars), scale))

```

```{r}
credit_model_lr_scaled <- glm(
  default ~
    total_income +
    total_dti +
    loan_amount +
    term +
    credit_history_years +
    delinq_2y +
    months_since_last_delinq +
    months_since_90d_late +
    num_historical_failed_to_pay +
    inquiries_last_12m +
    credit_utilization +
    total_credit_lines +
    open_credit_lines +
    emp_length +
    homeownership +
    application_type,
  data = credit_model_scaled,
  family = binomial()
)

```


```{r}
library(broom)

scaled_results <- tidy(credit_model_lr_scaled) |>
  filter(term != "(Intercept)") |>
  mutate(
    abs_effect = abs(estimate)
  ) |>
  arrange(desc(abs_effect))

top5_predictors <- scaled_results |>
  slice_head(n = 5)

top5_predictors

```

```{r}
# Extract and rank standardized coefficients
top5_plot_data <- tidy(credit_model_lr_scaled) |>
  filter(term != "(Intercept)") |>
  mutate(abs_effect = abs(estimate)) |>
  arrange(desc(abs_effect)) |>
  slice_head(n = 5) |>
  arrange(estimate)


standardized_plot <- ggplot(top5_plot_data, aes(x = reorder(term, estimate), y = estimate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 5 Standardized Predictors of Default",
    subtitle = "Effects shown per 1 standard deviation increase",
    x = "Predictor",
    y = "Standardized Log-Odds Effect"
  ) +
  theme_minimal(base_size = 14)

```

```{r}
ggsave("standard_plot.png", plot = standardized_plot, width = 7, height = 4, units = "in", dpi = 300)

```




