---
title: "credit_exploration"
output: html_document
---
# libraries
```{r}
library(shiny)
library(tidyverse)
library(glue)
library(DT)
library(lubridate)
library(broom)
```

```{r}
credit_raw <- read_csv("../data/loans_full_schema.csv") 

credit <- credit_raw |>
  mutate(
    default = if_else(
      loan_status %in% c("Charged Off", "Default", "Late (31-120 days)"),
      1L, 0L
    )
  )
#This model predicts probability of non-completion, not realized default.
```
#cleaning the data

Dropping a few things that leak future information and are not valid predictors.
```{r}
credit <- credit |>
  select(
    -emp_title,
    -installment,   # derived from loan_amount + term + rate
    -interest_rate, # outcome of underwriting, not input
    -grade,
    -sub_grade,
    -initial_listing_status,
    -disbursement_method
  )
```

Employment Length is categorical and shoes be converted to numeric
```{r}
credit <- credit |>
  mutate(
    emp_length = as.numeric(emp_length),
    emp_length = if_else(is.na(emp_length), 0 , emp_length)
  )
```

Lets make some other variables human readable
```{r}
#state
credit <- credit |>
  mutate(state = factor(state))

#Homeownership
credit <- credit |>
  mutate(homeownership = factor(homeownership))

#verified income
credit <- credit |>
  mutate(verified_income = factor(verified_income))

#application type
credit <- credit |>
  mutate(application_type = factor(application_type))

#loan purpose
credit <- credit |>
  mutate(loan_purpose = factor(loan_purpose))

```

Now lets clean up joint and individual loans
```{r}
credit <- credit |>
  mutate(
    total_income = if_else(
      application_type == "joint" & !is.na(annual_income_joint),
      annual_income_joint,
      annual_income
    ),
    total_dti = if_else(
      application_type == "joint" & !is.na(debt_to_income_joint),
      debt_to_income_joint,
      debt_to_income
    )
  )

#drop the raw components
credit <- credit |>
  select(
    -annual_income,
    -annual_income_joint,
    -debt_to_income,
    -debt_to_income_joint,
    -verification_income_joint
  )
```

Fixing the credit years variable
```{r}
credit <- credit |>
  mutate(
    # Ensure earliest_credit_line is numeric year
    earliest_credit_line = as.numeric(as.character(earliest_credit_line)),

    # Convert issue_month to date
    issue_date = as.Date(paste0("01-", issue_month), format = "%d-%b-%Y"),

    # Extract issue year
    issue_year = as.integer(format(issue_date, "%Y")),

    # Compute credit history length
    credit_history_years = issue_year - earliest_credit_line,

    # Guardrails
    credit_history_years = pmax(credit_history_years, 0),
    credit_history_years = pmin(credit_history_years, 50)
  ) |>
  select(
    -issue_month,
    -issue_date,
    -earliest_credit_line
  )

```


Now I want to create interpretable ratios
```{r}
credit <- credit |>
  mutate(
    credit_utilization = total_credit_utilized / total_credit_limit,
    debit_utilization  = total_debit_limit / pmax(num_active_debit_accounts, 1)
  )
```

Now I have to deal with missing values
```{r}
credit <- credit |>
  mutate(
    months_since_last_delinq = if_else(
      is.na(months_since_last_delinq), 999, months_since_last_delinq
    ),
    months_since_90d_late = if_else(
      is.na(months_since_90d_late), 999, months_since_90d_late
    )
  )

```

# Logistic Regression
```{r}
credit_model <- credit |>
  filter(
    loan_status %in% c(
      "Fully Paid",
      "Charged Off",
      "Late (31-120 days)"
    )
  )
```

create a target variable
```{r}
credit_model <- credit_model |>
  mutate(
    default = if_else(
      loan_status %in% c("Charged Off", "Late (31-120 days)"),
      1L, 0L
    )
  )

#dropping loan status because we labeled it
credit_model <- credit_model |>
  select(-loan_status)
```

```{r}
credit_model |>
  count(default) |>
  mutate(pct = n / sum(n))

```
Default rate somewhere in the 10–25% range

```{r}
table(credit$loan_status)
```
I excluded current loans because PD models require observed outcomes. Current and early-stage delinquent loans haven’t resolved yet, so including them would bias the target. I restricted training to finalized outcomes, which is standard practice in retail credit risk.

For a baseline PD model, we want:
-Credit behavior
-Capacity
-Stability
-Minimal redundancy
-No post-decision leakage

I needed to make sure all variables are usable in a model.
```{r}
credit_model <- credit_model |>
  mutate(
    total_income = as.numeric(total_income),
    total_dti = as.numeric(total_dti),
    loan_amount = as.numeric(loan_amount),
    credit_history_years = as.numeric(credit_history_years),
    delinq_2y = as.numeric(delinq_2y),
    months_since_last_delinq = as.numeric(months_since_last_delinq),
    months_since_90d_late = as.numeric(months_since_90d_late),
    num_historical_failed_to_pay = as.numeric(num_historical_failed_to_pay),
    inquiries_last_12m = as.numeric(inquiries_last_12m),
    credit_utilization = as.numeric(credit_utilization),
    total_credit_lines = as.numeric(total_credit_lines),
    open_credit_lines = as.numeric(open_credit_lines),
    emp_length = as.numeric(emp_length),

    # Explicit categorical casting
    term = factor(term),
    homeownership = factor(homeownership),
    application_type = factor(application_type)
  )

```


```{r}
credit_model_lr <- glm(
  default ~
    total_income +
    total_dti +
    loan_amount +
    term +
    credit_history_years +
    delinq_2y +
    months_since_last_delinq +
    months_since_90d_late +
    num_historical_failed_to_pay +
    inquiries_last_12m +
    credit_utilization +
    total_credit_lines +
    open_credit_lines +
    emp_length +
    homeownership +
    application_type,
  data = credit_model,
  family = binomial()
)


```


```{r}
coef_table <- tidy(credit_model_lr, conf.int = TRUE, exponentiate = TRUE)

coef_table |>
  arrange(desc(estimate))

```

# Model Performance

```{r}
credit_model_lr_results <- credit_model |>
  mutate(
    pd = predict(credit_model_lr, type = "response")
  )
summary(credit_model_lr_results)
```

```{r}
library(pROC)

roc_obj <- roc(
  response = credit_model_lr_results$default,
  predictor = credit_model_lr_results$pd
)

auc_value <- auc(roc_obj)
auc_value

```

```{r}
plot(
  roc_obj,
  col = "blue",
  main = paste0("ROC Curve (AUC = ", round(auc_value, 3), ")")
)
```
Explanation: 
- ROC shows how well the model ranks defaults above non-defaults
- AUC ≈ probability a random bad loan scores worse than a random good loan

0.5 = useless
0.6–0.7 = weak but usable
0.7–0.8 = solid credit model
0.8+ = very strong (rare for consumer credit)

```{r}
library(ggplot2)

ggplot(credit_model_lr_results, aes(x = pd, fill = factor(default))) +
  geom_histogram(
    bins = 50,
    position = "identity",
    alpha = 0.5
  ) +
  scale_fill_manual(
    values = c(
      "0" = "#003fba",  # No Default
      "1" = "#706363"   # Default
    )
  ) +
  labs(
    title = "Predicted PD Distribution by Loan Outcome",
    x = "Predicted Probability of Default",
    fill = "Default"
  )
```


```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    pd_decile = ntile(pd, 10)
  )

credit_model_lr_results |>
  group_by(pd_decile) |>
  summarise(
    avg_pd = mean(pd),
    default_rate = mean(default),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(pd_decile))

```

The model demonstrates strong ranking ability, with an AUC of X. Default rates increase monotonically across PD deciles, indicating effective risk separation.

# Risk Bands

```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    risk_band = case_when(
      pd < 0.05 ~ "Low",
      pd < 0.15 ~ "Medium",
      TRUE ~ "High"
    ),
    risk_band = factor(risk_band, levels = c("Low", "Medium", "High"))
  )

```

default risk bands
```{r}
credit_model_lr_results |>
  group_by(risk_band) |>
  summarise(
    avg_pd = mean(pd),
    observed_default_rate = mean(default),
    n = n(),
    .groups = "drop"
  )

```

visual of risk bands
```{r}
ggplot(credit_model_lr_results, aes(x = risk_band, fill = factor(default))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c("0" = "#003fba",
               "1" = "#706363") 
  ) +
  labs(
    title = "Observed Default Rate by Risk Band",
    x = "Risk Band",
    y = "Proportion",
    fill = "Default"
  )

```
decision logic

```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    decision = case_when(
      risk_band == "Low" ~ "Approve",
      risk_band == "Medium" ~ "Review",
      TRUE ~ "Decline"
    )
  )
credit_model_lr_results
```
Summary: 
Predicted PDs were translated into risk bands aligned with hypothetical credit policy thresholds. These bands allow consistent decisioning and enable transparent communication of risk to non-technical stakeholders.

# Lets explain somethings here

```{r}
coef_table <- tidy(
  credit_model_lr,
  conf.int = TRUE
)

coef_table <- coef_table |>
  mutate(
    estimate = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high)
  )

coef_table |>
  arrange(desc(estimate))


```

```{r}
coef_table_clean <- coef_table |>
  filter(term != "(Intercept)") |>
  mutate(
    direction = if_else(estimate > 1, "Increases Risk", "Decreases Risk"),
    pct_change_odds = (estimate - 1) * 100
  ) |>
  arrange(desc(abs(pct_change_odds)))
coef_table_clean
```

```{r}
#work to make sure this is working for each variable to explain what is making the prediction change
X <- model.matrix(credit_model_lr)
betas <- coef(credit_model_lr)

contrib <- sweep(X, 2, betas, `*`)

explain_one <- function(row_id) {
  tibble(
    feature = colnames(contrib),
    contribution = contrib[row_id, ]
  ) |>
    filter(feature != "(Intercept)") |>
    arrange(desc(abs(contribution)))
}

explain_one(1)
```

```{r}
explain_one(1) |>
  mutate(
    impact = case_when(
      contribution > 0 ~ "Increases PD",
      TRUE ~ "Decreases PD"
    )
  )

```

High credit utilization and renting status increase PD, while longer credit history reduces PD.

```{r}
print_individual_report <- function(row_id, model, data, risk_thresholds = c(Low = 0.05, Medium = 0.15), top_n = 3) {
  
  # Generate report
  report <- full_individual_report(row_id, model, data, risk_thresholds, top_n)
  
  cat("===========================================\n")
  cat("Individual Credit Report\n")
  cat("Observation ID:", report$row_id, "\n")
  cat("Predicted PD:", scales::percent(report$predicted_PD, accuracy = 0.1), "\n")
  cat("Risk Band:", report$risk_band, "\n")
  cat("===========================================\n\n")
  
  # Top contributors
  cat("Top", top_n, "Contributors to Predicted PD:\n")
  print(report$top_contributors |> 
          mutate(
            contribution_prob = scales::percent(contribution_prob, accuracy = 0.1)
          ) |>
          select(feature, value, contribution_prob, direction))
  cat("\n")
  
  # Variable effects
  cat("Effect of Increasing Each Variable by 1 Unit:\n")
  print(report$variable_effects |> 
          mutate(prob_change = scales::percent(prob_change, accuracy = 0.1)))
  cat("===========================================\n\n")
}

# Example usage
print_individual_report(1, credit_model_lr, credit_model_lr_results)

```



```{r}
credit_model_lr_results <- credit_model_lr_results |>
  mutate(
    decision_reason = case_when(
      risk_band == "High" ~ "High predicted default risk",
      risk_band == "Medium" ~ "Moderate risk — requires review",
      TRUE ~ "Low predicted default risk"
    )
  )

```
The model produces a Probability of Default, maps it to risk bands, and provides feature-level explanations for each decision, aligning with model risk management and fair lending expectations.


```{r}
all_vars <- names(coef(credit_model_lr))
all_vars
```
## Tibble for high risk factors

for logistic regression 
n=XB

impact of a variable
Impact(J) = B(J)(X(J)- X(J))

create an interactive table that generates impact of each variable based on the comparison with the average value and actual value.

```{r}
explain_table <- reactive({
  
  # Current borrower
  new_data <- borrower_data()
  
  # Model matrix for borrower
  X_new <- model.matrix(
    delete.response(terms(credit_model_lr)),
    new_data
  )
  
  # Model matrix for entire training data
  X_train <- model.matrix(
    delete.response(terms(credit_model_lr)),
    credit_model
  )
  
  # Average borrower (column means)
  X_avg <- colMeans(X_train)
  
  betas <- coef(credit_model_lr)
  
  results <- tibble(
    feature = colnames(X_new),
    impact = NA_real_
  )
  
  #FOR LOOP
  for (i in seq_along(results$feature)) {
    
    var_name <- results$feature[i]
    
    if (var_name %in% names(betas)) {
      
      borrower_value <- X_new[1, var_name]
      avg_value <- X_avg[var_name]
      beta_value <- betas[var_name]
      
      results$impact[i] <- beta_value * (borrower_value - avg_value)
    }
  }
  
  results |>
    filter(feature != "(Intercept)") |>
    arrange(desc(abs(impact))) |>
    slice_head(n = 5)
})

```











